---
title: "ASC_TIMEcourse_limma_nucleolus_analysis"
author: "Sarah Hp"
date: "6/4/22"
output: html_document
---

Calculate time effect for adipose stem cell differentiation to adipocytes via RNAseq. Here we extract the differentially expressed genes for Pro - D0, D0 - D3 and D3 - D15 for two donors.   Previously Pro, D0, D1, D3, and D9 were sequenced in triplicate for two independent donors (GSE176020).
New samples: day15 floating and bulk RNAseq.  


output files:
 - adipogenesis_rpkm_tmm.tab
 - adipogenesis_rpkm_tmm_means.tab
 - nucleolus_adipogenesis_DE.tab

intermediate files (merge of serveral experiments): 
 - late_adipo_and_D1&D2_native_rnaseq.counts
 - late_adipo_and_D1&D2_native_rnaseq_info.tab

```{r warning=F, message=F, echo=F}
#DE tools
library(limma)
library(edgeR)
library(biomaRt) #annotation
#plotting
library(ggplot2)
library(RUVSeq)
#GO
library(clusterProfiler)
library(org.Hs.eg.db)
#Data manipulation
library(tidyr)
library(dplyr)
```


```{r}
read.geo <- function(file_url) {
  con <- gzcon(url(file_url))
  txt <- readLines(con)
  return(read.delim(textConnection(txt), skip=1, header=T))
}
```

# Merge  experiments
Skip ahead to `Start analysis here` for DGE analysis from intermediate files.  

## Load Donor 2 timecourse from GEO
(with fractionally assigned multimapping alignments).  Then format the table. 

```{r}
d2_file = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE176020&format=file&file=GSE176020%5Fd13%5Frnaseq%2Emultim%2Ecounts%2Etxt%2Egz"
d2_tab = read.geo(d2_file)
colnames(d2_tab) =  gsub("output.01hisat.","", gsub(".sorted.bam","",colnames(d2_tab)))
#head(d2_tab) #table is messy so I won't print it here
```


Load file with sample info and match sample information to columns in FC table
```{r}
d2_info = read.delim("../sample_info/donor2_info.csv", header=TRUE, stringsAsFactors = FALSE, sep = ",")
#head(d2_info)

#remove unnecessary columns from d2_info
names(d2_info)[names(d2_info)=='Additional..fill.in.2.3.digits.code..max.5..construct.etc..'] = 'donor'
names(d2_info)[names(d2_info)=='Replicate..fill.in.number.'] = 'biorep'
names(d2_info)[names(d2_info)=='timept'] = 'time'
d2_info = d2_info[c('sample_id','time','donor','biorep','Tube.Label')]

#Checked that FC table and d2_info are in the same order
#colnames(d2_tab)
#d2_info$sample_id
```

## Load the Donor 1 timecourse: 
And generate sample information from the bam file names
```{r}
d1_file = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE176020&format=file&file=GSE176020%5Fmar19%5Frnaseq%2Emultim%2Ecounts%2Etxt%2Egz"
d1_tab = read.geo(d1_file)
colnames(d1_tab) =  gsub("output.01hisat.","", gsub(".sorted.bam","",colnames(d1_tab)))
head(d1_tab)

#Remove the file system artefacts from the count column names to create a list of the libraries:
sample_id = grep("[[:digit:]]",colnames(d1_tab), value = T)
sample_id
#Next create a sample table with information about samples and time points taken from the filenames.
Tube.Label = gsub("\\..*_S[[:digit:]]+", "", sample_id)
time = gsub(".*(KD|P)[43586]D?", "", gsub("_S.*","",sample_id))
time = factor(gsub("NT","-2",time)) #match factors
biorep = gsub(".*\\.(KD|P)", "P", gsub("(D[0139]|NT)_S.*","",sample_id))
biorep =gsub("P3","P8",biorep) #fix biorep error
donor = rep("D1G", 16)

d1_info = data.frame(sample_id,time,donor,biorep,Tube.Label)
d1_info
```

## Load New Samples (day 15)
These are from the same donors

```{r}
day15_tab = read.delim("../02featureCounts/day15_bulk_v_float.counts", skip=1,header=T, stringsAsFactors = F)
colnames(day15_tab) = gsub("output.01hisat.","", gsub(".sorted.bam","",colnames(day15_tab)))
colnames(day15_tab)
```

```{r}
day15_info = read.delim("../sample_info/day15_bulk_v_float_info.csv", header=T, sep=",")
day15_info$sample_id = gsub("-",".", day15_info$sample_id)
day15_info$time = factor(gsub("D","", day15_info$time))
colnames(day15_info) = gsub("rep", "biorep", colnames(day15_info))
#use cell_type column as experiment delimiter
day15_info$exp = 3

day15_info = day15_info[!(colnames(day15_info) %in% c("bio.condition"))]

colnames(day15_info); day15_info$sample_id
```
Edits to day15 file to fit: 
- remove D from time; add dash to donor
- change column names: rep to biorep; cell_type to exp
- remove researcher,  bio.condition

## Merge read counts and sample info
Merging the fc tables directly.  Merge sample information for d1 and d2; remove tube.label column, add exp column (since d1 and d2 have overlapping samples) and add separaion = bulk for these experiments.

```{r}
early = merge(d2_tab, d1_tab)
colnames(early); dim(early)

#Add experiment separator
d2_info$exp = 2
d1_info$exp = 1
colnames(d2_info)
colnames(d1_info)

early_info = rbind(d2_info, d1_info)
early_info$separation = "bulk"

early_info = early_info[colnames(early_info) != "Tube.Label"]

table(paste(early_info$time, early_info$donor, early_info$exp))
```


Merging in day15 now

```{r}
whole = merge(early, day15_tab)
dim(whole)

whole_info = rbind(early_info, day15_info)
str(whole_info); tail(whole_info)
```

Save combined read counts so we don't need to do this again. 

```{r}
write.table(whole, "../03limma/late_adipo_and_D1&D2_native_rnaseq.counts", sep="\t", row.names = F)
write.table(whole_info, "../03limma/late_adipo_and_D1&D2_native_rnaseq_info.tab", sep="\t", row.names = F)
```

# Start Analysis Here

Load combined read counts from above 
```{r read_in}
whole = read.delim("../03limma/late_adipo_and_D1&D2_native_rnaseq.counts", sep="\t", check.names = F )
whole_info = read.delim( "../03limma/late_adipo_and_D1&D2_native_rnaseq_info.tab", sep="\t")
```


## Make DGE object
remove additional samples.  

```{r}
whole_info$time.donor = paste(whole_info$time, whole_info$donor, sep=".")
whole_info$group = paste(whole_info$time.donor, whole_info$separation, sep=".")

whole_ob = DGEList(counts=data.matrix(whole[grep("[[:digit:]]", colnames(whole))]),  
                   genes= whole[c("Geneid", "Length")], samples = whole_info)
rownames(whole_ob$counts) = whole_ob$genes$Geneid
summary (whole_ob)
```

```{r message=FALSE}
#Check column names are the same
whole_ob$samples$sample_id
colnames(whole_ob)
```


## Initial Plots

Everything looks good. 1. MDSplot separates timepoints well 2. the RLE plot shows that median counts between libraries are 0, meaning library size and distribution is similar between libraries.  Therefore normalisation between experiments is not necessary.  
```{r message=FALSE}
plotMDS(whole_ob, labels = whole_ob$samples$time, gene.selection = "common")
plotRLE(whole_ob$counts, outline=FALSE, col=whole_ob$samples$group, main="Before filtering")
```
## Filtering

```{r}
plotSmear(whole_ob, main = "Before Filtering")
dim(whole_ob)

filt_series = whole_ob[filterByExpr(whole_ob),, keep.lib.sizes=FALSE]
dim(filt_series) #21174 genes kept (about 1k more genes that the timecourse to day9)
#The filter is approxiamtely 10/ libsize in millions
10/(median(whole_ob$samples$lib.size)/1000000) #0.245 minimum CPM 

#sanity check that gene_names match up
filt_series$counts[filt_series$genes$Geneid == "ENSG00000228630",] #HOTAIR
filt_series$counts[filt_series$genes$Geneid == "ENSG00000132170",] #PPARG

plotSmear(filt_series, main = "After Filtering") 
```

Donor vs donor this looks like very nice filtering (above, at any one time point some genes will be missing).  

```{r}
#Plot donor v donor -> a few low exression genes still but predominantly good filtering
two_filt = filt_series
two_filt$samples$group = as.factor(filt_series$samples$donor)
plotSmear(two_filt, main = "After Filtering")
```


```{r}
#library sizes are different, but relative log expression is
#now ~ normally distributed within libraries
plotRLE(filt_series$counts, outline=FALSE, 
        col=filt_series$samples$group, main="After filtering")
plotMDS(filt_series, labels = filt_series$samples$group, gene.selection = "common")
```

Because we're mainly interested in days Pro,0, 3 and 15 (bulk) I'm going to create a filter vector that eliminates genes that would drop out if we only used those samples.  The limma voom method requires adequate filtering of low counts genes to maintain conservative p-values.  

```{r}
short_ob = whole_ob[, whole_ob$samples$time %in% c(-2,0,3,15) &
                       whole_ob$samples$separation == "bulk"]
short_ob$samples
strict_filt_genes = short_ob[filterByExpr(short_ob),, keep.lib.sizes=FALSE]$genes
nrow(strict_filt_genes)
```

## Normalise
```{r}
filt_series = calcNormFactors(filt_series,method = "TMM")
filt_series$samples$norm.factors
plotRLE(cpm(filt_series$counts), outline=FALSE, 
        col=filt_series$samples$group, main="After TMM normalisation (no libsize)")
```

Library size is approximately the same after CPM normalisation, but it still needs to be adjusted directly for library size:

```{r}
plotRLE(cpm(filt_series, normalized.lib.sizes = TRUE), outline=FALSE, 
        col=filt_series$samples$group, main="After TMM normalisation (with libsize)")
```

## Annotate Gene Lists

```{r}
library(biomaRt)
mart <- biomaRt::useMart(biomart = "ensembl",
  dataset = "hsapiens_gene_ensembl",
   host = "https://jan2019.archive.ensembl.org")
annot = getBM(c("external_gene_name", "description", "ensembl_gene_id"), 
              filters = "ensembl_gene_id",
              values = filt_series$genes$Geneid,
              mart = mart)

head(annot, n=2); dim(annot)

#Tidying up the annot table
annot$description = gsub("\\[Source:.+\\]", "", annot$description)
colnames(annot)[1] = "gene_name"

#Add gene names to filt_series
filt_series$genes = merge(filt_series$genes, annot, by.x= "Geneid", 
                        by.y = "ensembl_gene_id", sort=FALSE)
#head(filt_series$genes)
#make sure length is numeric
filt_series$genes$Length = as.numeric(filt_series$genes$Length)
```

## Calculate FPKM

```{r}
rpkm = as.data.frame(rpkm(filt_series, normalized.lib.sizes = T))
rpkm$gene = rownames(rpkm)
colnames(rpkm) = paste("day",filt_series$samples$group, "_rep",filt_series$samples$biorep, sep="")
head(rpkm)

rpkm = rpkm[order(colnames(rpkm))]

format_rpkm = merge(filt_series$genes, rpkm,  
                    by.x="Geneid", by.y = 'row.names',sort=FALSE)
head(format_rpkm)
```

```{r}
write.table(format_rpkm, sep='\t', row.names = FALSE,quote = F,
            file="../03limma/adipogenesis_rpkm_tmm.tab")
```



```{r}
filt_series$samples$time.donor = filt_series$samples$group 
filt_series$samples$group = factor(paste("day",filt_series$samples$group, sep=""))
group_rpkm = data.frame(rpkmByGroup(filt_series,normalize.lib.sizes=TRUE))

colSums(group_rpkm)

format_grpkm = merge(filt_series$genes, group_rpkm,  
                    by.x="Geneid", by.y = 'row.names',sort=FALSE)
head(format_grpkm)
```

```{r}
rowMeans(format_grpkm[format_grpkm$gene_name == "PPARG",c(6:ncol(format_grpkm))])#expr matches the replicate table
plotRLE(data.matrix(format_grpkm[6:ncol(format_grpkm)]), outline=FALSE)
```

```{r}
write.table(format_grpkm, sep='\t', row.names = FALSE, quote=F,
            file="../03limma/adipogenesis_rpkm_tmm_means.tab")
```



## Limma Log Normalisation
```{r}
filt_series$samples$group = filt_series$samples$time.donor
norm_series = voom(filt_series, plot = TRUE)
head(norm_series$design, n=2) #groups are time.donor
summary(norm_series$E[,1])#log normlisation min -7, max 15
```


## Design

```{r}
#1 check the default design
colnames(norm_series$design)
head(filt_series$samples, n=2)

pair_design = model.matrix(~0+filt_series$samples$group)
colnames(pair_design) = paste("day", gsub("-",".",levels(filt_series$samples$group)), sep="")
rownames(pair_design) = filt_series$samples$sample_id
head(pair_design, n=2)
```

## Making a large time based model
Using bulk day15 samples
time_effect_donor1 + time_effect_donor2 was the pvalue used at 0.01 to select genes for dpgp

```{r}
pair_fit = lmFit(norm_series, design=pair_design)

contrasts = makeContrasts(time_effect_donor1 = (day0.D1G.bulk - day.2.D1G.bulk ) + #D-2 -> D0 effect
                                                  (day1.D1G.bulk - day0.D1G.bulk) + #D0 -> D1 effect
                                                  (day3.D1G.bulk - day1.D1G.bulk) + #D1 -> D3 effect
                                                  (day9.D1G.bulk - day3.D1G.bulk) + #D3 -> D9 effect 
                                                  (day15.D1G.bulk - day9.D1G.bulk),
                          time_effect_donor2 = (day0.D2A.bulk - day.2.D2A.bulk ) + #D-2 -> D0 effect
                                                  (day1.D2A.bulk - day0.D2A.bulk) + #D0 -> D1 effect
                                                  (day3.D2A.bulk - day1.D2A.bulk) + #D1 -> D3 effect
                                                  (day9.D2A.bulk - day3.D2A.bulk) + #D3 -> D9 effect 
                                                  (day15.D2A.bulk - day9.D2A.bulk), #D9 -> D15 effect
                          simple_time_effect = (day0.D1G.bulk - day.2.D1G.bulk ) + #D-2 -> D0 effect
                                                  (day3.D1G.bulk - day0.D1G.bulk) + #D0 -> D3 effect
                                                  (day15.D1G.bulk - day3.D1G.bulk) + #D3 -> D15 effect
                                                  (day0.D2A.bulk - day.2.D2A.bulk ) + #D-2 -> D0 effect
                                                  (day3.D2A.bulk - day0.D2A.bulk) + #D1 -> D3 effect
                                                  (day15.D2A.bulk - day3.D2A.bulk), #D9 -> D15 effect
                          #Donor 7 changes over time
                          d.2_to_d0.D1 = day0.D1G.bulk - day.2.D1G.bulk,
                          d0_to_d1.D1 = day1.D1G.bulk - day0.D1G.bulk,
                          d1_to_d3.D1 = day3.D1G.bulk - day1.D1G.bulk, 
                          d3_to_d9.D1= day9.D1G.bulk - day3.D1G.bulk,
                          d9_to_d15.D1.bulk =  day15.D1G.bulk - day9.D1G.bulk,
                          d9_to_d15.D1.floating =  day15.D1G.floating - day9.D1G.bulk,
                          d15.bulk_to_floating.D1 = day15.D1G.floating - day15.D1G.bulk,
                          #Donor 13 changes over time
                          d.2_to_d0.D2 = day0.D2A.bulk - day.2.D2A.bulk,
                          d0_to_d1.D2 = day1.D2A.bulk - day0.D2A.bulk,
                          d1_to_d3.D2 = day3.D2A.bulk - day1.D2A.bulk,
                          d3_to_d9.D2 = day9.D2A.bulk - day3.D2A.bulk,
                          d9_to_d15.D2.bulk =  day15.D2A.bulk - day9.D2A.bulk,
                          d9_to_d15.D2.floating =  day15.D2A.floating - day9.D2A.bulk,
                          d15.bulk_to_floating.D2 = day15.D2A.floating - day15.D2A.bulk,
                          #And the additional timepoints
                          d0_to_d3.D1 =  day3.D1G.bulk - day0.D1G.bulk,
                          d0_to_d3.D2 =  day3.D2A.bulk - day0.D2A.bulk,
                          d3_to_d15.D1 =  day15.D1G.bulk - day3.D1G.bulk,
                          d3_to_d15.D2 =  day15.D2A.bulk - day3.D2A.bulk,
                          levels=pair_design)
special_fit = contrasts.fit(pair_fit,contrasts)
special_fit = eBayes(special_fit, robust = TRUE)
```

```{r}
summary(decideTests(special_fit, method="separate", adjust.method = "BH", p.value = 0.01)) #Adjusted for contrast
summary(decideTests(special_fit ,method="global", adjust.method = "BH"), p.value= 0.01) #Unadjusted for contrast
```

Interesting, the time effect is less significant in donor2.  The time effect is similar rather than different between donors -> most genes respond similarly to the adipogenic cocktail, regardless of depot/donor.  Removing those extra timepoints (Day1, Day9), leads to ~6k genes found both up and down

## Number of DE genes

Adding an extra timepoint adds more genes to the DE list; 11,639 genes at p < 0.01 (basically 50% of all detected genes).  

```{r}
comb = topTable(special_fit, number = nrow(pair_fit$genes), coef = c("time_effect_donor1","time_effect_donor2"))
dim(comb[comb$adj.P.Val < 0.05,]) #More than half of all genes change; most will be the same
dim(comb[comb$adj.P.Val < 0.01,]) 

volcanoplot(special_fit, coef = c(1,2), highlight =10, names = special_fit$genes$gene_name,
            main="Time effect (combined across donors)", xlab="log2FC ABDO <---> GLUT")

volcanoplot(special_fit, coef = c(1,2), highlight =11639, names = special_fit$genes$gene_name,
            main="Time effect (combined across donors)", xlab="log2FC ABDO <---> GLUT")

volcanoplot(special_fit, coef = c(1,2),
            main="Time effect (combined across donors)", xlab="log2FC ABDO <---> GLUT")

head(comb)
```

```{r}
nuc = topTable(special_fit, number = nrow(pair_fit$genes), coef = "simple_time_effect")
dim(nuc[nuc$adj.P.Val < 0.05,]) #More than half of all genes change; most will be the same
dim(nuc[nuc$adj.P.Val < 0.01,]) 

nuc = nuc[nuc$Geneid %in% strict_filt_genes$Geneid,] #We can filter out ~450 genes with low expression but "significant" at these timepoints
nrow(nuc)

head(nuc)
```


```{r eval=F}
write.table(nuc,  sep="\t", row.names=FALSE, 
            file.path("../03limma/nucleolus_adipogenesis_DE.tab"))
```


```{r}
ggplot(nuc) + geom_histogram(aes(x=P.Value), bins=200)
volcanoplot(special_fit, coef="simple_time_effect", highlight = 12, names = special_fit$genes$gene_name)
```
## Venn diagram of new and old lists
```{r}
res = decideTests(special_fit, method="separate", adjust.method = "BH", p.value = 0.01) #Adjusted for contrast

vennDiagram(res[,c("simple_time_effect","time_effect_donor1", "time_effect_donor2")], include="up")

vennDiagram(res[,c("simple_time_effect","time_effect_donor1", "time_effect_donor2")], include="down")
```

Okay, so there's about 500 unique genes is either direction discovered by this test (as opposed to the d1->d3->d9 tests).  We lose about the same number from either donor, but none that were in common between the donors, which seems promising.  It seems that we capture the bulk of the adipogenic changes with just timepoints Pro, 0, 3 and 15 :)

