---
title: "ribosome_heatmaps_adipogenesis"
author: "Sarah Hp"
date: '2022-07-08'
output: pdf_document
---

Changes:

- Using DE calculated from only 4 time points (Pro -> D0 -> D3 -> D15)

From previous revisions:

- printing lists of gene to annotate the terms/heatmaps

```{r}
library(biomaRt)
library(ComplexHeatmap)
knitr::opts_chunk$set(echo = TRUE, dev = c("pdf"), fig.path = "ribosome_heatmaps_adipogenesis/", fig.dim = c(4, 8))
```


```{r}
rpkm = read.delim( "../03limma/adipogenesis_rpkm_tmm_means.tab", header=T) #rpkm table
head(rpkm); dim(rpkm)

combpval = read.delim("../03limma/nucleolus_adipogenesis_DE.tab", header=T)
head(combpval)

sig = merge(rpkm, combpval[c("gene_name","adj.P.Val")])
head(sig); dim(sig)
```


```{r}
#formatting
sig = sig[sig$adj.P.Val < 0.01,]
sig$adj.P.Val = NULL
#discard duplicate rownames
sig = sig[!duplicated(sig$gene_name),]
dim(sig)

rownames(sig) = sig$gene_name
sig$gene_name = NULL
```

remove floating adipocytes
```{r}
sig = sig[!grepl("floating", colnames(sig))]
head(sig)

sig = sig[grepl("day(.2|0|3|15)", colnames(sig))]
#separate by donor
d1 = sig[grep("D1G", colnames(sig), value=T)]
head(d1)
d2 = sig[grep("D2A", colnames(sig), value=T)]
head(d2)
```



## Get GO terms

```{r}
mart <- biomaRt::useMart(biomart = "ensembl",
  dataset = "hsapiens_gene_ensembl",
  host = "https://jan2019.archive.ensembl.org") 

cyt_ribosome = biomaRt::getBM(c("external_gene_name", "ensembl_gene_id","go_linkage_type"), 
              filters = "go",
              values = c("GO:0022625","GO:0022627"),
              mart = mart)
length(unique(cyt_ribosome$ensembl_gene_id)) #120 cytosolic ribosome genes
```


```{r}
translation = getBM(c("external_gene_name", "ensembl_gene_id"), 
              filters = "go",
              values = "GO:0006412",
              mart = mart)
nrow(translation)

norp_trans = translation[!grepl("^M?RP", translation$external_gene_name),]
nrow(norp_trans) #165 non RP translation genes
```

```{r}
ribogen = biomaRt::getBM(c("external_gene_name", "ensembl_gene_id","go_linkage_type"), 
              filters = "go",
              values = c("GO:0042254"),
              mart = mart)
length(unique(ribogen$ensembl_gene_id)) #104

norp_ribogen = ribogen[!grepl("^M?RP", ribogen$external_gene_name),]
length(unique(norp_ribogen$ensembl_gene_id)) #96
```

Check the gene lists
```{r}
summary(translation$external_gene_name %in% rownames(sig))
summary(ribogen$external_gene_name %in% rownames(sig))
```


# Donor 1

## heatmap formatting
```{r}
colnames(d1) = gsub(".D1G.bulk","", colnames(d1))
d1 = d1[c("day.2","day0","day3","day15")]

#create zscores of log transformed rpkms
zs = t(scale(t(log2(d1+1))))
 # before log scaling the median is always about -0.2, meaning theres a negative bias to the values
summary(zs) # the median is more variable between timepoints upon log scaling

#create vector for ordering
#order = as.vector(zs[,"day15"] - zs[,"day.2"])
order = apply(zs, 1, cor, y=1:ncol(zs)) #uses pearson correlation
head(order)
```

Cytosolic ribosome components with significantly different expression cross the timecourse.  
```{r d1_ribo}
summary(rownames(zs) %in% cyt_ribosome$external_gene_name) #72 genes

Heatmap(as.matrix(zs[rownames(zs) %in% cyt_ribosome$external_gene_name,]),
        cluster_columns = F,
        name="Relative FPKM", column_title = "Cytosolic ribosome",
        col=circlize::colorRamp2(c(-2,0,2),c(rgb(0.2,0.3,0.6),rgb(1,1,1),rgb(0.9,0.1,0.1))),
         row_names_gp = gpar(fontsize = 7),
         show_row_names = FALSE, 
        row_dend_reorder = order[rownames(zs) %in% cyt_ribosome$external_gene_name])

```

## Translation terms


And with splits

```{r trans_split}
Heatmap(as.matrix(zs[rownames(zs) %in% translation$external_gene_name,]),
        cluster_columns = F,
        name="Relative FPKM",
        column_title = "Translation",
        col=circlize::colorRamp2(c(-2,0,2),c(rgb(0.2,0.6,1),rgb(0,0,0),rgb(0.95,0.3,0.3))),
        show_row_names = F, row_split=4)

Heatmap(as.matrix(zs[rownames(zs) %in% translation$external_gene_name,]),
        cluster_columns = F,
        name="Relative FPKM",
        column_title = "Translation",
        col=circlize::colorRamp2(c(-2,0,2),c(rgb(0.2,0.6,1),rgb(0,0,0),rgb(0.95,0.3,0.3))),
        show_row_names = F, row_split=8)


```

### Figure3a Lower panel

```{r Figure3a_lower}
Heatmap(as.matrix(zs[rownames(zs) %in% translation$external_gene_name,]),
        cluster_columns = F,
        name="Relative FPKM",
        column_title = "Translation",
        col=circlize::colorRamp2(c(-2,0,2),c(rgb(0.2,0.6,1),rgb(0,0,0),rgb(0.95,0.3,0.3))),
        show_row_names = T, row_split=4, row_names_gp = gpar(fontsize = 4),
        row_dend_reorder = order[rownames(zs) %in% translation$external_gene_name])
```



## Ribosome biogenesis


```{r ribogen_split}

Heatmap(as.matrix(zs[rownames(zs) %in% ribogen$external_gene_name,]),
        cluster_columns = F,
        name="Relative FPKM",
        column_title = "Ribosome biogenesis",
        height=0.25*length(unique(translation$external_gene_name)),
        width=0.5*ncol(d1),
         col=circlize::colorRamp2(c(-2,0,2),c(rgb(0.2,0.6,1),rgb(0,0,0),rgb(0.95,0.3,0.3))),
         row_names_gp = gpar(fontsize = 8), row_split=6,
        row_dend_reorder = order[rownames(zs) %in% ribogen$external_gene_name])


```

### Figure3a Upper panel
```{r Figure3a_upper}
Heatmap(as.matrix(zs[rownames(zs) %in% ribogen$external_gene_name,]),
        cluster_columns = F,
        name="Relative FPKM",
        column_title = "Ribosome biogenesis",
        height=0.25*length(unique(translation$external_gene_name)),
        width=0.5*ncol(d1),
         col=circlize::colorRamp2(c(-2,0,2),c(rgb(0.2,0.6,1),rgb(0,0,0),rgb(0.95,0.3,0.3))),
         row_names_gp = gpar(fontsize = 8), row_split=3,
        row_dend_reorder = order[rownames(zs) %in% ribogen$external_gene_name])

```


# Donor 2

## heatmap formatting
```{r}
colnames(d2) = gsub(".D2A.bulk","", colnames(d2))
d2 = d2[c("day.2","day0","day3","day15")]

#create zscores
zs13 = t(scale(t(log2(d2+1))))
head(zs13)

# create order vector
#order13 = as.vector(zs13[,"day15"] - zs13[,"day.2"])
order13 = apply(zs13, 1, cor, y=1:ncol(zs13))
```

Cytosolic ribosome components with significantly different expression cross the timecourse.  Day 9 and floating adipocytes share high expression of many of these genes.  
```{r d2_ribo}
summary(rownames(zs13) %in% cyt_ribosome$external_gene_name) #72 genes

Heatmap(as.matrix(zs13[rownames(zs13) %in% cyt_ribosome$external_gene_name,]),
        cluster_columns = F,
        name="Relative FPKM", column_title = "Cytosolic ribosome",
        height=0.5*length(unique(cyt_ribosome$external_gene_name)),
        width=0.5*ncol(d1),
        col=circlize::colorRamp2(c(-2,0,2),c(rgb(0.2,0.3,0.6),rgb(1,1,1),rgb(0.9,0.1,0.1))),
        row_dend_reorder = order13[rownames(zs13) %in% cyt_ribosome$external_gene_name],
        show_row_names = F)
```

## D2 Translation Heatmaps

```{r d2_trans_split}
Heatmap(as.matrix(zs13[rownames(zs13) %in% translation$external_gene_name,]),
        cluster_columns = F,
        name="Relative FPKM",
        column_title = "Translation",
        col=circlize::colorRamp2(c(-2,0,2),c(rgb(0.2,0.6,1),rgb(0,0,0),rgb(0.95,0.3,0.3)))  ,
        show_row_names = F, row_split = 4,
        row_dend_reorder = order13[rownames(zs13) %in% translation$external_gene_name])


```

### Supp Figure 6a lower panel

```{r FigureS6a_lower}
Heatmap(as.matrix(zs13[rownames(zs13) %in% translation$external_gene_name,]),
        cluster_columns = F,
        name="Relative FPKM",
        column_title = "Translation",
        col=circlize::colorRamp2(c(-2,0,2),c(rgb(0.2,0.6,1),rgb(0,0,0),rgb(0.95,0.3,0.3)))  ,
        show_row_names = T, row_split = 4, row_names_gp = gpar(fontsize=4),
        row_dend_reorder = order13[rownames(zs13) %in% translation$external_gene_name])
```

## D2 Ribosome Biogenesis Heatmaps

```{r d2_ribogen_split}
Heatmap(as.matrix(zs13[rownames(zs13) %in% ribogen$external_gene_name,]),
        cluster_columns = F,
        name="Relative FPKM",
        column_title = "Ribosome Biogenesis",
        col=circlize::colorRamp2(c(-2,0,2),c(rgb(0.2,0.6,1),rgb(0,0,0),rgb(0.95,0.3,0.3))),
        show_row_names = T, row_names_gp = gpar(fontsize=8),
        row_dend_reorder = order13[rownames(zs13) %in% ribogen$external_gene_name], 
        row_split=4)

Heatmap(as.matrix(zs13[rownames(zs13) %in% ribogen$external_gene_name,]),
        cluster_columns = F,
        name="Relative FPKM",
        column_title = "Ribosome Biogenesis",
        col=circlize::colorRamp2(c(-2,0,2),c(rgb(0.2,0.6,1),rgb(0,0,0),rgb(0.95,0.3,0.3))),
        show_row_names = T, row_names_gp = gpar(fontsize=8),
        row_dend_reorder = order13[rownames(zs13) %in% ribogen$external_gene_name], 
        row_split=8)
```

### Supp Figure 6a upper panel

```{r FigureS6a_upper}
Heatmap(as.matrix(zs13[rownames(zs13) %in% ribogen$external_gene_name,]),
        cluster_columns = F,
        name="Relative FPKM",
        column_title = "Ribosome Biogenesis",
        col=circlize::colorRamp2(c(-2,0,2),c(rgb(0.2,0.6,1),rgb(0,0,0),rgb(0.95,0.3,0.3))),
        show_row_names = T, row_names_gp = gpar(fontsize=8),
        row_dend_reorder = order13[rownames(zs13) %in% ribogen$external_gene_name], 
        row_split=2)
```

